---
title: "EU-MOHP v013.1.0 Dataset"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', fig.pos = "H", out.width = "100%", message = FALSE, warning = FALSE, fig.showtext = TRUE)
options(knitr.table.format = 'markdown')
library(targets)
library(tidyverse)
```

The EU-MOHP v013.1.0 dataset was developed and generated as part of the [MACRO project](https://www.bgr.bund.de/EN/Themen/Wasser/Projekte/laufend/F+E/Macro/macro_projektbeschr_en.html?nn=1548270).

This repository hosts the dataset EU-MOHP v013.1.0 as single GeoTIFF files with .tif file extension but compressed to *.7z* files for download.
The corresponding publication, which is linked in the corresponding [github repository](https://github.com/MxNl/macro_mohp_feature), provides all background information. It also explains the file naming scheme. To select the files relevant to your study case or application, please use the search bar to filter by spatial coverage or specific hydrologic orders by entering one of the values listed in the table below. Note, however, that not every region or spatial coverage applies to all hydrologic orders. If you want to check in more detail whether your area of interest is covered by this dataset and which files are relevant, please use this [interactive map](https://mxnl.github.io/macro_mohp_feature/).

<br />

```{r outputdata, echo=FALSE}
tabledata <-
  targets::tar_read(output_data_table) %>%
  mutate(`Placeholder in output file name` = 
           str_remove_all(`Placeholder in output file name`, "<|>")) %>%
  rename(`Value (use this in search bar)` = Value) %>%
  mutate(row = row_number(), .before = 1)

placegolder_one <-
  tabledata %>%
  filter(`Placeholder in output file name` == "region name for spatial coverage") %>%
  pull(row) %>%
  range()
placegolder_two <-
  tabledata %>%
  filter(`Placeholder in output file name` == "abbreviation of the EU-MOHP measure") %>%
  pull(row) %>%
  range()
placegolder_three <-
  tabledata %>%
  filter(`Placeholder in output file name` == "hydrologic order") %>%
  pull(row) %>%
  range()

tabledata %>%
  select(-row) %>%
  knitr::kable(
    # booktabs = TRUE,
  ) %>%
  # kableExtra::kable_styling(font_size = 12) %>%
  kableExtra::row_spec(placegolder_one[1]:placegolder_one[2], background = "#f2f2f2") %>%
  kableExtra::row_spec(placegolder_two[1]:placegolder_two[2], background = "white") %>%
  kableExtra::row_spec(placegolder_three[1]:placegolder_three[2], background = "#f2f2f2")
# kableExtra::kable_styling(full_width = TRUE) %>%
# kableExtra::column_spec(2, width = "12em") %>%
# kableExtra::collapse_rows()
```

<br />

The actively developed code can be found in this [github repository](https://github.com/MxNl/macro_mohp_feature).

