---
title: "Multiorder Hydrologic Position in Europe as a Set of Metrics in Support of Groundwater Mapping at Regional and National Scales"
author:
  - name: Maximilian Noelscher
    affiliation:
    - Federal Institute for Geosciences and Natural Resources (BGR), Berlin, Germany
abstract: |
  This dataset (EU-MOHP v013.0.1) provides information on the multiorder hydrologic position of a geographic point within its respective river network or catchment. More precisely, it comprises the three measures “lateral position” as a relative measure of the position between the stream and the catchment boundary/ watershed, “divide stream distance” as an absolute distance measure that serves as a proxy for the position within the catchment and “stream distance” as an absolute measure of the distance to the nearest stream. These three measures were calculated for several hydrologic (stream) orders. Its spatial extent covers major parts of physiographical Europe and all of the 39 countries in European Economic Area (EEA39). Although there might be many potential use cases, this dataset serves predominantly as valuable input data for mapping tasks in the context of hydrogeology and subsurface characteristics in general. \cite{belitz_multiorder_2019}
output: 
  bookdown::pdf_document2:
    number_sections: FALSE
    template: scienctific_data_template.tex
    keep_tex: yes
bibliography: 
  - eumohp.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center', fig.pos = "H")
options(tinytex.verbose = TRUE)

library(DiagrammeR)
library(DiagrammeRsvg)
library(rsvg)
library(knitr)

source("../targets/R/visualization_functions.R")
```

```{r, echo=FALSE}
options(encoding = "UTF-8")
RefManageR::ReadZotero("5574385",
               .params = list(collection = "U4DY76U3",
                              key = "V0zS3yc7zT1NsT1bfHG6sfYQ")) %>%
      bibtex::write.bib("eumohp.bib")
# 
# renv::dependencies()$Package %>%
#       purrr::chuck(1) %>%
#       # knitr::write_bib() %>% 
#       bibtex::write.bib("packages.bib", append = TRUE)
```


<!--
```{r eval=FALSE, code=readLines("../targets/R/visualization_functions.R")}
 
```
-->


```{r, echo=FALSE}
make_workflow_diagram("diagramms/flowchart.pdf")
```


# Background & Summary

In recent years, data science tools such as machine learning are increasingly applied to and specifically developed for hydro(geo)logical challenges and research questions. In the field of hydrogeology, machine learning has been used successfully for groundwater level prediction and a variety of mapping  tasks. Since machine learning models are traditionally based purely on data with no built-in knowledge of physical processes, it is important to provide as many variables (predictor variables/ explanatory variables/ features) as possible that have an impact on the target variable to potentially enable the machine learning algorithm to reproduce the result of the underlying  process. For surface and near-surface processes, this criterion may be more or less satisfiable through the availability of remote sensing data, whereas for modelling subsurface processes such as in hydrogeology, this poses a serious challenge. \cite{desimone_machine-learning_2020, belitz_multiorder_2019}

```{r}
asd <- "sdf"
```

# Methods {-}

All processing and analysis was conducted with free open source software. All processing steps except for the data download that was done manually are controlled and executed from within a targets pipeline in the programming language R [!!source]. Targets is an R package that provides a toolkit for reproducible workflows [!!source]. Spatial vector data such as the !!rivers are processed partly in R and a PostgreSQL database (version 13) with a PostGIS (version 3.1.0) extension for speed and memory reasons. For the same reason, all major raster calculations were conducted in a GRASS GIS database (version 7.8.5-2). The database connections and all calculations in the databases are also controlled by the targets pipeline. For reaching a maximum of reproducibility, a docker container is provided to rerun all calculations with little effort. The R package renv is used for keeping track of the required R package versions and combines well with targets and docker to endure reproducibility.

## Detailed Workflow {-}

In the following, the description of the methods is oriented towards the structure of the targets pipeline to easily relate the methods description here to the source code in the repository. All steps required to understand the workflow will be described, for further details we refer to the source code.

### Step 1: Data Acquisition {-}
The “EU-Hydro - River Network Database” was manually downloaded from https://land.copernicus.eu (for detailed link see references) as version v013. All downloaded and unzipped files have approximately 14 GB. The !!river is the only underlying data for the generation of the EU-MOHP dataset.

## Hardware {-}
The pipeline to generate the dataset was executed on a DELL PowerEdge C4140 Server with an Intel Xeon Gold 6240R CPU and 384 GB installed RAM. The installed operation system is Microsoft Windows Server 2019 Standard, version 10.0.17763 Build 17763.

[what is different to the Beelitz Paper and why]
NHDPlusV2 data
No pathleveId column
Criterion to exclusively use free open source software

\@ref(fig:flowchart)

# Data Records {-}

Text.

# Technical Validation {-}

Text.

# Usage Notes {-}

Text.

# Code availability {-}

Text.

# Acknowledgements {-}

Text.

# Author contributions statement {-}

Text.

# Competing interests {-}

Text.

# Figures \& Tables {-}

```{r flowchart, echo=FALSE, out.width = "70%", fig.cap= "sdf"}
include_graphics("diagramms/flowchart.pdf") 
```
